Thesis Title

### 2022-02-24 chapters/introduction.tex
add corrections from grammarly

The Global Burden of Disease (GBD) 2019 study reported that microbial pathogens are responsible for more than 400 million years of life lost annually across the globe, a higher burden than either cancer or cardiovascular disease \citep{vos_global_2020}. In particular, lower respiratory infections, diarrhoeal diseases, HIV/AIDS and tuberculosis were amongst the five leading causes of of global total years of life lost. More recently, the COVID-19 pandemic, declared as such by the World Health Organization (WHO) on 11 March 2020 after the emergence and global spread of the severe acute respiratory syndrome coronavirus 2 (SARS-CoV-2), and as of January 2022 has cause more than 5.63 million deaths worldwide \citep{ritchie_coronavirus_2020}, making it one of the deadliest pandemics in history. Coronavirus have been responsible for three of the eighteen major pandemics registered throughout modern history \citep{piret_pandemics_2021}, all occurring after the year 2000. \textit{Yersinia pestis}, responsible for three pandemics of plague, \textit{Vibrio cholerae}, with seven cholera pandemics, and Influenza A virus, the causative agent of five flu pandemics, are responsible for the remaining ones, with Influenza being the only other pathogen with a pandemic registered after 2000. Recent decades have also witnessed the emergence of additional virulent pathogens, including the Ebola virus, West Nile virus, Dengue virus and Zika virus, particularly in lower-income countries.

@@

Clinical microbiology is a discipline focused on rapidly characterising pathogen samples in order to direct the management of individual infected patients (diagnostic microbiology) and monitor the epidemiology of infectious disease (public health microbiology), including the detection of outbreaks and infection prevention. According to WHO's Global Expenditure on Health report from 2000 to 2019, of the 51 countries that reported health spending by disease and condition, an average of 37\% of health spending went to infectious and parasitic diseases, corresponding to the largest share of health spending \citep{world_health_organization_global_2021}. About 21\% of total health spending went to three major infectious diseases — HIV/AIDS (9\%), tuberculosis (1\%) and malaria (11\%) — and 16\% went to other infectious and parasitic diseases. On average, 70\% of external aid for health went to infectious and parasitic diseases in the 51 low and middle income countries. Of the \$54.8 billion estimated disbursed for health in 2020, \$13.7 billion (25\%) was targeted toward the COVID-19 health response \citep{micah_tracking_2021}.

@@

The past few decades have seen a major revolution in the operation of microbial laboratories, driven by the development of molecular technologies and ways to make these accessible, namely amplification-based polymerase chain reaction (PCR), matrix-assisted laser desorption/ionisation - time of light (MALDI-TOF) and DNA-microarray-based hybridisation technology. These are used in conjunction with traditional techniques such as microscopy, culture and serology, not fully replacing them. Application of these methods differs by suspected infection type: bacterial, viral, fungal or parasitic.

@@

The current gold standard for bacterial pathogen identification in diagnostic microbiology laboratories involves the isolation of the pathogen through culture followed by biochemical testing, multi-step process that can take days to weeks before obtaining results, depending on the fastidiousness of the organism and if it can be cultured \citep{muhamad_rizal_advantages_2020, giuliano_guide_2019, muhamad_rizal_advantages_2020}. Although culture allows the identification of a wide variety of organisms, some pathogens can escape routine investigation due to strict metabolic necessities for growth or requirement for specific biochemical tests needed for their identification. Additionally, results will be obscured if a mixed culture is obtained, particularly if the cultures are obtained from sites with a microbiota, such as the gut and the skin, increasing the risk of contamination from normal flora, leading to false results \citep{giuliano_guide_2019}. After a successful growth in culture, Gram staining and MALDI-TOF mass spectometry is often used for identification with good accuracy as long as the pathogen is presented in the the coexisting database \citep{patel_maldi-tof_2015}. An alternate rapid identification method is PCR where nucleic acid fragments are detected through specific probes, being highly sensitive and specific, to the point where PCR may detect bacteria that are not viable after a patient has been treated for an infection \citep{scerbo_beyond_2016}. It's limited to the probes used, with multiplex PCR, an extension of PCR by using multiple primers, tried to address this issue by allowing for the identification of multiple bacteria and other important information such as the detection of antibiotic resistance genes or virulence genes \citep{giuliano_guide_2019}

Following identification, antibiotic-susceptibility testing is essential for guiding clinicians in selecting an appropriate treatment. Conventional detection methods of bacterial resistance, such as disc diffusion, antimicrobial gradient strip and broth microdilution, are widely used but results cannot be obtained earlier than 48 hours after receiving a sample, which may lead to prolonged use or overuse of broad-spectrum antibiotics \citep{benkova_antimicrobial_2020}. Similarly to bacterial identification, MALDI-TOF and PCR has been increasingly adopted as solutions with lower turnaround time, although no phenotipic information is retrieved, nor information on the minimum inhibitory concentration (MIC) for a given antibiotic.

Choosing an appropriate bacterial typing technique for epidemiological studies depends on the resources available and the minimum intended resolution, ranging from DNA fingerprinting to multilocus sequence typing, Pulsed-field gel electrophoresis and Sequence-based typing \citep{allerberger_molecular_2012,foxman_choosing_2005}. DNA macrorestriction analysis by pulsed field gel electrophoresis (PFGE), which revolutionised precise separation of DNA fragments, became the most widely implemented DNA fingerprinting technique \citep{allerberger_molecular_2012}, becoming the golden standard for bacterial typing \citep{neoh_pulsed-field_2019} .

In the early 2000s, Multilocus sequence typing (MLST) was proposed as a portable, universal, and definitive method for characterising bacteria \citep{maiden_multilocus_2006}. Instead of enzyme restriction of bacteria DNA, separation of the restricted DNA bands using a pulsed-field electrophoresis chamber, followed by clonal assignment of bacteria based on PFGE banding patterns, MLST relies on the amplification through PCR sequences of internal fragments of house-keeping genes (usually 5 to 7), approximately 450-500 basepairs (bp) in size, followed by its the sequence, usually my Sanger methods. For each house-keeping gene, the different sequences present within a bacterial species are assigned as distinct alleles and, for each isolate, the alleles at each of the (usually) seven loci define the allelic profile or sequence type \citep{larsen_multilocus_2012}. As with PFGE, different schemes, defining what house-keeping gene fragments are used, are available depending of the species. Unlike PFGE, the provision of freely accessible, curated databases of MLST nucleotide sequence data enables the direct comparison of bacterial isolates, providing the basis of a common language for bacterial typing \citep{maiden_multilocus_2006}. So far, MLST schemes for 115 bacterial organisms having been published and made freely available (\url{https://pubmlst.org/organisms}, \cite{jolley_open-access_2018})

@@

The traditional approaches to laboratory diagnosis of viral infections have been (1) direct detection in patient material of virions, viral antigens, or viral nucleic acids, (2) isolation of virus in cultured cells, followed by identification of the isolate, and (3) detection and measurement of antibodies in the patient’s serum (serology) \citep{burrell_laboratory_2017}. Viral diagnostics is therefore generally organised into two primary categories, indirect and direct detection, depending on the method used.

Indirect detection methods involve the propagation of virus particles via their introduction to a suitable host cell line (virus isolation), as viruses rely on host organisms to replicate. This is a relatively slow diagnostic method, sometimes taking weeks for the virus to propagate, usually followed by miscroscopy for its identification, or more commonly, through molecular methods with an agent which detects a virus-associated protein, such as an antibody \citep{cassedy_virus_2021}.

Direct detection methods negate the need for virus propagation, detecting the virus directly from the suspect source through nucleic acid and immunological methods. PCR and reverse transcription-PCR are widely applied methods for the detection of both DNA and RNA viruses, respectively, driven by an increased awareness of the clinical value of, and demand for, prompt information about viral loads, viral sequence data, and potential antiviral resistance information \citep{cassedy_virus_2021}. Syndromic testing, utilising multiplex PCR to simultaneously amplify the nucleic acid of multiple targets in a single reaction, are now fully integrated into the standard testing practices of many clinical laboratories \citep{dien_bard_panels_2020}. Limitations of these assays include no detection of off-target pathogens, a lack of full susceptibility information, cost, and false-positive results.

Real-time quantitative PCR (qPCR) remains the front line tool in aetiological diagnosis, measuring the production of the target amplicon throughout the reaction and providing quantitative results with high specificity and sensibility, albeit with a significant cost due to sophisticated apparatus despite high-throughput systems being widely established \citep{cassedy_virus_2021}. Immunoassays employ singular-epitope specificity antibodies as the primary means to detect viruses within a sample, and provides as a much more cost-efficient alternative to nucleic acid detection \citep{cassedy_virus_2021}. One major application is seroprevalence assays, an essential technique for identifying patients who have been exposed to a virus (historical exposure), detect asymptomatic infection or evaluate vaccine efficacy  \citep{chan_determining_2021, bobrovitz_global_2021}. Lateral flow immunoassays (LFIA) are extensively used for detecting virus-associated protein directly from source through labelled antibodies binding to their cognate antigens, usually read by way of a colour change at a test line. Besides being very cost effective, LFAIs have a turnaround time of minutes and the colour change can be observed with the naked eye, therefore facilitating rapid diagnosis but its results are limited to semi-quantitative and it does not typically achieve sensitivity comparable to nucleic-acid detection \citep{estrela_lateral_2016, cassedy_virus_2021, di_nardo_ten_2021}.

@@

Invasive fungal infections are a prevalent disease in pulmonary and critical care medicine, and they frequently pose diagnostic challenges, with laboratory diagnosis of fungal infections being identified as a key area in need of development of a clinical practise guideline \citep{haydour_diagnosis_2019}. Both serologic and molecular approaches are currently in use, with the first including not only immunodiffusion and complement fixation, both of which remain a cornerstone for fungal diagnostic testing, but also lateral-flow assays for fungal antigen detection. Molecular techniques for fungal identification both from culture and directly from patient specimens, include nucleic acid probes, mass spectrometry-based methods, and nucleic acid amplification testing \citep{ramanan_laboratory_2017}. Classic fungal culture continues to be relevant and necessary, with most detection and identification methods requiring growth in culture, from phenotypic identification and staining techniques, to nucleic acid hybridisation probes and MALDI-TOF mass spectrometry.

Molecular methods of fungal identification directly from the suspect source are particularly attractive for fungi that fail to produce characteristic morphologic characters in a timely fashion, but molecular databases for fungi tend to lag behind bacterial databases in terms of maturity \citep{ramanan_laboratory_2017}. qPCR testing  is one of the most frequently used
molecular methods in clinical microbiology, even days before the onset of positive culture result \citep{kourkoumpetis_polymerase_2012}. The ribossomal RNA (rRNA) gene cluster is frequently used as target for fungi identification, as well as other targets to evaluate antifungal resistance, guiding therapy, which lowers patient mortality and decrease unnecessary antifungal treatment, improving treatment-associated costs and avoiding toxicity \citep{kourkoumpetis_polymerase_2012}.

@@

Currently, the detection and diagnosis of parasite infections rely on several laboratory methods in addition to clinical symptoms, clinical history, travel history, and geographic location of patient. serology-based assays, molecular-based assays and microscopy  \citep{ndao_diagnosis_2009}. The use of the microscopy for the morphology-based diagnosis of disease remains the reference standard in parasitology although identification and matching of the life-cycle stages, despite its low cost, is very challenging even with expert knowledge \citep{blasco-costa_molecular_2016}. LFAIs has been used successfully, overcoming many of the disadvantages of conventional microscopy by providing simple, quick, cost-effective, and sensitive and accurate identification of parasites, particularly in low-income countries, but are available for a relatively low number of organisms, low sensitivity and reports of cross-reactivity \citep{momcilovic_rapid_2019, robert-gangneux_epidemiology_2012, rajput_false_2018}.

@@

Infectious disease surveillance is critical for improving population health, generating information that drives action not only in the management of infected patients in the but also in the prevention of new ones by identifying emerging health conditions that may have a significant impact by (1) describing the current burden and epidemiology of disease, (2) to monitoring trends, and (3) identifying outbreaks and new pathogens \citep{groseclose_public_2017, murray_infectious_2017}. Public health surveillance systems (PHSS) are composed by the ongoing systematic collection, analysis, and interpretation of data, and its integration with the timely dissemination of results to those who can undertake effective prevention and control activities \citep{teutsch_considerations_2010}.

Traditional PHSS can have different approaches based on the epidemiology and clinical presentation of the disease and the goals of surveillance. In passive surveillance systems, medical professionals in the community and at health facilities report cases to the public health agency, which conducts data management and analysis once the data are received and communicate with the responsible entities. Globally, the WHO as describes in the International Health Regulations what is notifiable by every country to WHO, such as Severe acute respiratory syndrome (SARS) and Viral haemorrhagic fevers
(Ebola, Lassa, Marburg), as well as providing guidance on what public health measures should be implemented \citep{world_health_organization_international_2005}. Active surveillance aims to detect every case, nor relying on a reporting structure, and can have many approaches from sentinel sites or network of sites that capture cases of a given condition, such as respiratory tract infections, within a catchment population \citep{murray_infectious_2017, melo-cristino_estudo_2006}. The application of environmental surveillance methods, performed prospectively to detect pathogens prior to the recording of clinical cases or to monitor their abundance in environment to assess the potential risk of disease, has been proven as a viable alternative particularly in wastewater \citep{andrews_environmental_2020, mcweeney_demonstration_1894, baker_combined_2011, larsen_tracking_2020}.

"One Health" is an collaborative and multi-disciplinary approach to designing and implementing programmes, policies, legislation and research in which multiple sectors communicate and work together to achieve better public health outcomes \citep{mackenzie_one_2019}. It recognises that people’s health is closely connected to animals’ health and shared environment, focusing on zoonotic and vector-borne diseases, antimicrobial resistance, food safety, food security and environmental contamination \citep{rugarabamu_one-health_2021}.

@@

Since the publication of the first complete microbial genome a quarter of a century ago, that of the bacterium \textit{Haemophilus influenzae} \citep{hood_dna_1996}, genomics has transformed the field of microbiology, and in particular its clinical application. The paper describing the DNA-sequencing method with chain-terminating inhibitors used in the sequencing of the first microbial genome \citep{sanger_dna_1977}, which earned the late Frederick Sanger his share of the 1980 Nobel Prize in Chemistry alongside Walter Gilbert, was, in 2014, the top fourth in number of citations with 60335, highlighting its impact in the field of biological sciences, and by extension medicine \citep{van_noorden_top_2014}. Currently, this number has increased to 84546 according to PubMed Central\textsuperscript{\small\textregistered} (PMC) (\url{https://pubmed.ncbi.nlm.nih.gov/, https://www.ncbi.nlm.nih.gov/pmc/articles/PMC431765/}). Since its emergence, reductions in cost, technical advances in sequencing technologies and new computational developments have made genomic sequencing one of the most influential tools in biomedical research, yielding unprecedented insights into microbial evolution and diversity, and the complexity of the genetic variation in both commensal and pathogenic microbes.

@@

Since the  the discovery of the structure of DNA \citep{watson_molecular_1953}, great strides have been made in understanding the complexity and diversity of genomes in health and disease. The ability to determine the sequence in a DNA molecule  
The development and commercialisation of high-throughput, massively parallel sequencing, has democratised sequencing by offering individual laboratories, either in research or in health, access to the technology.

\subsubsection{The first generation of DNA sequencing} \label{ssec:1st_gen_seq}

In the late 1980s, automated Sanger sequencing machines could sequence approximately 1,000 bases per day, having been applied in the 1990s to large bacterial genomes and the first unicellular and multicellular eukaryotic genomes, including the completion of a high-quality, reference sequence of the human genome under the Human Genome Project (HGP) \citep{koch_sequencing_2021, collins_human_1995}. The first genomes of the pathogenic \textit{Mycobacterium tuberculosis} \citep{cole_deciphering_1998}, \textit{Yersinia pestis} \citep{parkhill_genome_2001}, \textit{Escherichia coli} K-12 \citep{blattner_complete_1997} were sequenced using this technology, requiring years of effort and significant budgets but providing insights into the genomic complexity of these organisms. Some of the complete genome sequences produced during this era are still used today as high quality references.

In a simplistic manner, in Sanger sequencing, also known as “first-generation” DNA sequencing, a DNA polymerase is used to synthesize numerous copies of the sequence of interest using dideoxynucleotide triphosphates (ddNTPs) spiked into the reaction. At each nucleotide incorporation event, there is chance that a ddNTP will be added and the growing DNA chain will be terminated, resulting in a collection of DNA molecules of varying lengths \citep{sanger_dna_1977, hagemann_overview_2015}. Modern Sanger sequencing uses fluorescently labelled ddNTPs that allow the amplification step to be performed in a single reaction, resulting in a mixture of single-stranded DNA fragments of various lengths, each tagged at one end with a fluorophore indicating the identity of the 3' nucleotide that, after separation through capillary electrophoresis, the resulting electropherogram with four-color fluorescence intensity can be interpreted by a base-calling software and producing 600–1000 bases of accurate sequence (reads) \citep{hagemann_overview_2015}.

The Sanger sequencing technology remains very useful for applications where high throughput is not required due to its cost effectiveness, relatively low sample load and accuracy of sequencing even in repetitive genomic regions, although input DNA must consist of a relatively pure population of sequences \citep{slatko_overview_2018}. One of the most common uses is thus individual sequencing reactions using a specific DNA primer on a specific template, such as MLST of bacterial genomes.

\subsubsection{The second generation of DNA sequencing} \label{ssec:2nd_gen_seq}

The release of the first truly high-throughput sequencing platform in the mid-2000s heralded a 50,000-fold drop in the cost of DNA sequencing in comparison with the first generation technologies and led to the denomination of next-generation sequencing (NGS) \citep{goodwin_coming_2016}. This trend has continued to throughout the next two decades of continued development and improvement, allied to the emergence of benchtop sequencing platforms with high throughput of sequencing data and turnaround times of days, making it a standard in any microbiology and public health laboratories \citep{loman_twenty_2015}. Second generation sequencing methods can be grouped into two major categories: (1) sequencing by hybridisation and (2) sequencing by synthesis.

>>>>

The Global Burden of Disease (GBD) 2019 study reported that microbial pathogens are responsible for more than 400 million years of life lost annually across the globe, a higher burden than either cancer or cardiovascular disease \citep{vos_global_2020}. In particular, lower respiratory infections, diarrhoeal diseases, HIV/AIDS and tuberculosis were amongst the five leading causes of global total years of life lost. More recently, the COVID-19 pandemic, declared as such by the World Health Organization (WHO) on 11 March 2020 after the emergence and global spread of the severe acute respiratory syndrome coronavirus 2 (SARS-CoV-2), and as of January 2022 has caused more than 5.63 million deaths worldwide \citep{ritchie_coronavirus_2020}, making it one of the deadliest pandemics in history. Coronavirus has been responsible for three of the eighteen major pandemics registered throughout modern history \citep{piret_pandemics_2021}, all occurring after the year 2000. \textit{Yersinia pestis}, responsible for three pandemics of plague, \textit{Vibrio cholerae}, with seven cholera pandemics, and Influenza A virus, the causative agent of five flu pandemics, are responsible for the remaining ones, with Influenza being the only other pathogen with a pandemic registered after 2000. Recent decades have also witnessed the emergence of additional virulent pathogens, including the Ebola virus, West Nile virus, Dengue virus and Zika virus, particularly in lower-income countries.

@@

Clinical microbiology is a discipline focused on rapidly characterising pathogen samples to direct the management of individual infected patients (diagnostic microbiology) and monitor the epidemiology of infectious disease (public health microbiology), including the detection of outbreaks and infection prevention. According to WHO's Global Expenditure on Health report from 2000 to 2019, of the 51 countries that reported health spending by disease and condition, an average of 37\% of health spending went to infectious and parasitic diseases, corresponding to the largest share of health spending \citep{world_health_organization_global_2021}. About 21\% of total health spending went to three major infectious diseases — HIV/AIDS (9\%), tuberculosis (1\%) and malaria (11\%) — and 16\% went to other infectious and parasitic diseases. On average, 70\% of external aid for health went to infectious and parasitic diseases in the 51 low and middle-income countries. Of the \$54.8 billion estimated disbursed for health in 2020, \$13.7 billion (25\%) was targeted toward the COVID-19 health response \citep{micah_tracking_2021}.

@@

The past few decades have seen a major revolution in the operation of microbial laboratories, driven by the development of molecular technologies and ways to make these accessible, namely amplification-based polymerase chain reaction (PCR), matrix-assisted laser desorption/ionisation - time of flight (MALDI-TOF) and DNA-microarray-based hybridisation technology. These are used in conjunction with traditional techniques such as microscopy, culture and serology, not fully replacing them. Application of these methods differs by suspected infection type: bacterial, viral, fungal or parasitic.

@@

The current gold standard for bacterial pathogen identification in diagnostic microbiology laboratories involves the isolation of the pathogen through culture followed by biochemical testing, a multi-step process that can take days to weeks before obtaining results, depending on the fastidiousness of the organism and if it can be cultured \citep{muhamad_rizal_advantages_2020, giuliano_guide_2019, muhamad_rizal_advantages_2020}. Although culture allows the identification of a wide variety of organisms, some pathogens can escape routine investigation due to strict metabolic necessities for growth or the requirement for specific biochemical tests needed for their identification. Additionally, results will be obscured if a mixed culture is obtained, particularly if the cultures are obtained from sites with a microbiota, such as the gut and the skin, increasing the risk of contamination from normal flora, leading to false results \citep{giuliano_guide_2019}. After successful growth in culture, Gram staining and MALDI-TOF mass spectrometry are often used for identification with good accuracy as long as the pathogen is presented in the the coexisting database \citep{patel_maldi-tof_2015}. An alternate rapid identification method is PCR where nucleic acid fragments are detected through specific probes, being highly sensitive and specific, to the point where PCR may detect bacteria that are not viable after a patient has been treated for an infection \citep{scerbo_beyond_2016}. It's limited to the probes used, with multiplex PCR, an extension of PCR by using multiple primers, tried to address this issue by allowing for the identification of multiple bacteria and other important information such as the detection of antibiotic resistance genes or virulence genes \citep{giuliano_guide_2019}

Following identification, antibiotic-susceptibility testing is essential for guiding clinicians in selecting an appropriate treatment. Conventional detection methods of bacterial resistance, such as disc diffusion, antimicrobial gradient strip and broth microdilution, are widely used but results cannot be obtained earlier than 48 hours after receiving a sample, which may lead to prolonged use or overuse of broad-spectrum antibiotics \citep{benkova_antimicrobial_2020}. Similarly to bacterial identification, MALDI-TOF and PCR have been increasingly adopted as solutions with lower turnaround times, although no phenotypic information is retrieved, nor information on the minimum inhibitory concentration (MIC) for a given antibiotic.

Choosing an appropriate bacterial typing technique for epidemiological studies depends on the resources available and the minimum intended resolution, ranging from DNA fingerprinting to multilocus sequence typing, Pulsed-field gel electrophoresis and Sequence-based typing \citep{allerberger_molecular_2012,foxman_choosing_2005}. DNA macrorestriction analysis by pulsed-field gel electrophoresis (PFGE), which revolutionised precise separation of DNA fragments, became the most widely implemented DNA fingerprinting technique \citep{allerberger_molecular_2012}, becoming the golden standard for bacterial typing \citep{neoh_pulsed-field_2019}.

In the early 2000s, Multilocus sequence typing (MLST) was proposed as a portable, universal, and definitive method for characterising bacteria \citep{maiden_multilocus_2006}. Instead of enzyme restriction of bacteria DNA, separation of the restricted DNA bands using a pulsed-field electrophoresis chamber, followed by clonal assignment of bacteria based on PFGE banding patterns, MLST relies on the amplification through PCR sequences of internal fragments of housekeeping genes (usually 5 to 7), approximately 450-500 basepairs (bp) in size, followed by its the sequence, usually my Sanger methods. For each house-keeping gene, the different sequences present within a bacterial species are assigned as distinct alleles and, for each isolate, the alleles at each of the (usually) seven loci define the allelic profile or sequence type \citep{larsen_multilocus_2012}. As with PFGE, different schemes, defining what house-keeping gene fragments are used, are available depending on the species. Unlike PFGE, the provision of freely accessible, curated databases of MLST nucleotide sequence data enables the direct comparison of bacterial isolates, providing the basis of a common language for bacterial typing \citep{maiden_multilocus_2006}. So far, MLST schemes for 115 bacterial organisms have been published and made freely available (\url{https://pubmlst.org/organisms}, \cite{jolley_open-access_2018})

@@

The traditional approaches to the laboratory diagnosis of viral infections have been (1) direct detection in patient material of virions, viral antigens, or viral nucleic acids, (2) isolation of the virus in cultured cells, followed by identification of the isolate, and (3) detection and measurement of antibodies in the patient’s serum (serology) \citep{burrell_laboratory_2017}. Viral diagnostics is therefore generally organised into two primary categories, indirect and direct detection, depending on the method used.

Indirect detection methods involve the propagation of virus particles via their introduction to a suitable host cell line (virus isolation), as viruses rely on host organisms to replicate. This is a relatively slow diagnostic method, sometimes taking weeks for the virus to propagate, usually followed by microscopy for its identification, or more commonly, through molecular methods with an agent which detects a virus-associated protein, such as an antibody \citep{cassedy_virus_2021}.

Direct detection methods negate the need for virus propagation, detecting the virus directly from the suspect source through nucleic acid and immunological methods. PCR and reverse transcription-PCR are widely applied methods for the detection of both DNA and RNA viruses, respectively, driven by increased awareness of the clinical value of, and demand for, prompt information about viral loads, viral sequence data, and potential antiviral resistance information \citep{cassedy_virus_2021}. Syndromic testing, utilising multiplex PCR to simultaneously amplify the nucleic acid of multiple targets in a single reaction, is now fully integrated into the standard testing practices of many clinical laboratories \citep{dien_bard_panels_2020}. Limitations of these assays include no detection of off-target pathogens, a lack of full susceptibility information, cost, and false-positive results.

Real-time quantitative PCR (qPCR) remains the front line tool in aetiological diagnosis, measuring the production of the target amplicon throughout the reaction and providing quantitative results with high specificity and sensibility, albeit with a significant cost due to sophisticated apparatus despite high-throughput systems being widely established \citep{cassedy_virus_2021}. Immunoassays employ singular-epitope specificity antibodies as the primary means to detect viruses within a sample and provide a much more cost-efficient alternative to nucleic acid detection \citep{cassedy_virus_2021}. One major application is seroprevalence assays, an essential technique for identifying patients who have been exposed to a virus (historical exposure), detecting asymptomatic infection or evaluating vaccine efficacy  \citep{chan_determining_2021, bobrovitz_global_2021}. Lateral flow immunoassays (LFIA) are extensively used for detecting virus-associated protein directly from the source through labelled antibodies binding to their cognate antigens, usually read by way of a colour change at a test line. Besides being very cost-effective, LFAIs have a turnaround time of minutes and the colour change can be observed with the naked eye, therefore facilitating rapid diagnosis but its results are limited to semi-quantitative and it does not typically achieve sensitivity comparable to nucleic-acid detection \citep{estrela_lateral_2016, cassedy_virus_2021, di_nardo_ten_2021}.

@@

Invasive fungal infections are a prevalent disease in pulmonary and critical care medicine, and they frequently pose diagnostic challenges, with laboratory diagnosis of fungal infections being identified as a key area in need of development of a clinical practise guideline \citep{haydour_diagnosis_2019}. Both serologic and molecular approaches are currently in use, with the first including not only immunodiffusion and complement fixation, both of which remain a cornerstone for fungal diagnostic testing, but also lateral-flow assays for fungal antigen detection. Molecular techniques for fungal identification both from culture and directly from patient specimens include nucleic acid probes, mass spectrometry-based methods, and nucleic acid amplification testing \citep{ramanan_laboratory_2017}. Classic fungal culture continues to be relevant and necessary, with most detection and identification methods requiring growth in culture, from phenotypic identification and staining techniques to nucleic acid hybridisation probes and MALDI-TOF mass spectrometry.

Molecular methods of fungal identification directly from the suspect source are particularly attractive for fungi that fail to produce characteristic morphologic characters in a timely fashion, but molecular databases for fungi tend to lag behind bacterial databases in terms of maturity \citep{ramanan_laboratory_2017}. qPCR testing is one of the most frequently used
molecular methods in clinical microbiology, even days before the onset of positive culture result \citep{kourkoumpetis_polymerase_2012}. The ribosomal RNA (rRNA) gene cluster is frequently used as the target for fungi identification, as well as other targets to evaluate antifungal resistance, guiding therapy, which lowers patient mortality and decreases unnecessary antifungal treatment, improving treatment-associated costs and avoiding toxicity \citep{kourkoumpetis_polymerase_2012}.

@@

Currently, the detection and diagnosis of parasite infections rely on several laboratory methods in addition to clinical symptoms, clinical history, travel history, and geographic location of the patient. serology-based assays, molecular-based assays and microscopy  \citep{ndao_diagnosis_2009}. The use of microscopy for the morphology-based diagnosis of the disease remains the reference standard in parasitology although identification and matching of the life-cycle stages, despite its low cost, is very challenging even with expert knowledge \citep{blasco-costa_molecular_2016}. LFAIs has been used successfully, overcoming many of the disadvantages of conventional microscopy by providing simple, quick, cost-effective, sensitive and accurate identification of parasites, particularly in low-income countries, but are available for a relatively low number of organisms, low sensitivity and reports of cross-reactivity \citep{momcilovic_rapid_2019, robert-gangneux_epidemiology_2012, rajput_false_2018}.

@@

Infectious disease surveillance is critical for improving population health, generating information that drives action not only in the management of infected patients but also in the prevention of new ones by identifying emerging health conditions that may have a significant impact by (1) describing the current burden and epidemiology of the disease, (2) to monitoring trends, and (3) identifying outbreaks and new pathogens \citep{groseclose_public_2017, murray_infectious_2017}. Public health surveillance systems (PHSS) are composed of the ongoing systematic collection, analysis, and interpretation of data, and its integration with the timely dissemination of results to those who can undertake effective prevention and control activities \citep{teutsch_considerations_2010}.

Traditional PHSS can have different approaches based on the epidemiology and clinical presentation of the disease and the goals of surveillance. In passive surveillance systems, medical professionals in the community and at health facilities report cases to the public health agency, which conducts data management and analysis once the data are received and communicate with the responsible entities. Globally, the WHO as described in the International Health Regulations what is notifiable by every country to WHO, such as Severe acute respiratory syndrome (SARS) and Viral haemorrhagic fevers
(Ebola, Lassa, Marburg), as well as guiding what public health measures should be implemented \citep{world_health_organization_international_2005}. Active surveillance aims to detect every case, not relying on a reporting structure, and can have many approaches from sentinel sites or network of sites that capture cases of a given condition, such as respiratory tract infections, within a catchment population \citep{murray_infectious_2017, melo-cristino_estudo_2006}. The application of environmental surveillance methods, performed prospectively to detect pathogens prior to the recording of clinical cases or to monitor their abundance in the environment to assess the potential risk of disease, has been proven as a viable alternative, particularly in wastewater \citep{andrews_environmental_2020, mcweeney_demonstration_1894, baker_combined_2011, larsen_tracking_2020}.

"One Health" is a collaborative and multi-disciplinary approach to designing and implementing programmes, policies, legislation and research in which multiple sectors communicate and work together to achieve better public health outcomes \citep{mackenzie_one_2019}. It recognises that people’s health is closely connected to animals’ health and shared environment, focusing on zoonotic and vector-borne diseases, antimicrobial resistance, food safety, food security and environmental contamination \citep{rugarabamu_one-health_2021}.

@@

Since the publication of the first complete microbial genome a quarter of a century ago, that of the bacterium \textit{Haemophilus influenzae} \citep{hood_dna_1996}, genomics has transformed the field of microbiology, and in particular its clinical application. The paper describing the DNA-sequencing method with chain-terminating inhibitors used in the sequencing of the first microbial genome \citep{sanger_dna_1977}, which earned the late Frederick Sanger his share of the 1980 Nobel Prize in Chemistry alongside Walter Gilbert, was, in 2014, the top fourth in the number of citations with 60335, highlighting its impact in the field of biological sciences, and by extension medicine \citep{van_noorden_top_2014}. Currently, this number has increased to 84546 according to PubMed Central\textsuperscript{\small\textregistered} (PMC) (\url{https://pubmed.ncbi.nlm.nih.gov/, https://www.ncbi.nlm.nih.gov/pmc/articles/PMC431765/}). Since its emergence, reductions in cost, technical advances in sequencing technologies and new computational developments have made genomic sequencing one of the most influential tools in biomedical research, yielding unprecedented insights into microbial evolution and diversity, and the complexity of the genetic variation in both commensal and pathogenic microbes.

@@

Since thes discovery of the structure of DNA \citep{watson_molecular_1953}, great strides have been made in understanding the complexity and diversity of genomes in health and disease. The ability to determine the sequence in a DNA molecule  
The development and commercialisation of high-throughput, massively parallel sequencing, has democratised sequencing by offering individual laboratories, either in research or in health, access to the technology.

\subsubsection{The first-generation of DNA sequencing} \label{ssec:1st_gen_seq}

In the late 1980s, automated Sanger sequencing machines could sequence approximately 1,000 bases per day, having been applied in the 1990s to large bacterial genomes and the first unicellular and multicellular eukaryotic genomes, including the completion of a high-quality, reference sequence of the human genome under the Human Genome Project (HGP) \citep{koch_sequencing_2021, collins_human_1995}. The first genomes of the pathogenic \textit{Mycobacterium tuberculosis} \citep{cole_deciphering_1998}, \textit{Yersinia pestis} \citep{parkhill_genome_2001}, \textit{Escherichia coli} K-12 \citep{blattner_complete_1997} were sequenced using this technology, requiring years of effort and significant budgets but providing insights into the genomic complexity of these organisms. Some of the complete genome sequences produced during this era are still used today as high-quality references.

Simplistically, in Sanger sequencing, also known as “first-generation” DNA sequencing, a DNA polymerase is used to synthesize numerous copies of the sequence of interest using dideoxynucleotide triphosphates (ddNTPs) spiked into the reaction. At each nucleotide incorporation event, there is a chance that a ddNTP will be added and the growing DNA chain will be terminated, resulting in a collection of DNA molecules of varying lengths \citep{sanger_dna_1977, hagemann_overview_2015}. Modern Sanger sequencing uses fluorescently labelled ddNTPs that allow the amplification step to be performed in a single reaction, resulting in a mixture of single-stranded DNA fragments of various lengths, each tagged at one end with a fluorophore indicating the identity of the 3' nucleotide that, after separation through capillary electrophoresis, the resulting electropherogram with four-colour fluorescence intensity can be interpreted by a base-calling software and producing 600–1000 bases of accurate sequence (reads) \citep{hagemann_overview_2015}.

The Sanger sequencing technology remains very useful for applications where high-throughput is not required due to its cost-effectiveness, relatively low sample load and accuracy of sequencing even in repetitive genomic regions, although input DNA must consist of a relatively pure population of sequences \citep{slatko_overview_2018}. One of the most common uses is thus individual sequencing reactions using a specific DNA primer on a specific template, such as MLST of bacterial genomes.

\subsubsection{The second-generation of DNA sequencing} \label{ssec:2nd_gen_seq}

The release of the first truly high-throughput sequencing platform in the mid-2000s heralded a 50,000-fold drop in the cost of DNA sequencing in comparison with the first-generation technologies and led to the denomination of next-generation sequencing (NGS) \citep{goodwin_coming_2016}. This trend has continued throughout the next two decades of continued development and improvement, allied to the emergence of benchtop sequencing platforms with a high-throughput of sequencing data and turnaround times of days, making it a standard in any microbiology and public health laboratories \citep{loman_twenty_2015}. second-generation sequencing methods can be grouped into two major categories: (1) sequencing by hybridisation and (2) sequencing by synthesis.
